Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-06-15 article-models-tt.tex
terms as sections
starting to remove the empty context

\subsubsection{The empty context}
\label{empty}
In the theory of contexts, families and terms we introduce an empty context and,
for every context $\Gamma$, an empty family over $\Gamma$. 
When contexts are viewed as statements, the empty context is the statement that
asserts nothing. Likewise, in any context one can assert nothing and this is
achieved by the empty family. 
Since somebody who states nothing asserts it all right, there is a term of the empty
family over any context. Moreover, any two terms of the empty family are always
judgmentally equal.
\begin{align}
& \inference
  { }
  { \jctx{\emptyc}
    }
  \\
& \inference
  { \jctx{\Gamma}
    }
  { \jfam{\Gamma}{\emptyf[\Gamma]}
    }
  \\
& \inference
  { \jctx{\Gamma}
    }
  { \jterm{\Gamma}{\emptyf[\Gamma]}{\emptytm[\Gamma]}
    }
  \\
& \inference
  { \jterm{\Gamma}{\emptyf[\Gamma]}{x}
    }
  { \jtermeq
      {\Gamma}
      {\emptyf[\Gamma]}
      {x}
      {\emptytm[\Gamma]}
    }
  \label{comp-00-t}
\end{align}

\label{comp-00}
Because every context is a family of contexts over the empty context, 
the empty context $\emptyc$ may be considered as a family of contexts over the empty
context. When we do this, we get $\emptyf[\emptyc]$.
\begin{equation}\label{eq:comp-00}
\inference
  { }
  { \jfameq
      {\emptyc}
      {\emptyc}
      {\emptyf[\emptyc]}
    }
\end{equation}
In the future, we shall denote $\emptyf[\Gamma]$ by $\emptyf$. The above rule
guarantees that this will not cause confusion. Likewise, we shall denote
$\emptytm[\Gamma]$ by $\emptytm$.

@@

In the following set of inference rules we state that the empty context and
the empty family are neutral objects when extension is concerned.
\label{comp-e0}\label{comp-0e}
\begin{align}
%& \inference
%  { \jctx{\Gamma}
%    }
%  { \jctxeq{\ctxext{\Gamma}{\emptyf}}{\Gamma}
%    }
%  \label{comp-e0-c}\\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{{A}{\emptyf}}{A}
    }
  \label{comp-e0-f}
  \\
%& \inference
%  { \jctx{\Gamma}
%    }
%  { \jctxeq{\ctxext{\emptyc}{\Gamma}}{\Gamma}
%    }
%  \label{comp-0e-c}\\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{\ctxext{\emptyf}{A}}{A}
    }
  \label{comp-0e-f}
\end{align}
Note the special case where we take $\Gamma$ to be the empty context.

@@

@@

\subsection{Further judgmental equalities of the theory of contexts, families and
terms}
We have already seen that empty families, extension, weakening, substitution
and the identity terms are compatible with each other. There are, however, a
few more judgmental equalities explaining what happens when
\begin{enumerate}
\item something is weakend by the empty family. This is used to consider a term
      $x$ of $A$ in context $\Gamma$ as a morphism from $\emptyf$ to $A$ in
      context $\Gamma$, see \autoref{morphisms};
\item something is substituted by the term of the empty family;
\item something is weakened by an extended family. This is Currying for weakening;
\item something is first weakened by a family $A$ in context $\Gamma$ and then
      substituted by a term of $A$. This is used to show that composition is
      compatible with itself (i.e. that composition is associative);
\item the identity terms are joining the party. Identity terms are empty contexts
      in certain models, so they should behave accordingly.
\end{enumerate}
In other words, the rules seem to not fit in the scheme that all operations are
compatible with each other, but in fact each of these rules plays a role in
completing the scheme (work out further).

>>>>

@@

@@

\subsubsection{The empty families}
\label{empty}
We introduce an empty family over $\Gamma$ for every context $\Gamma$. It was
a suggestion of Voevodsky and Garner to only include empty families and not an
empty context. Having an empty context requires also rules asserting that a
context is the same thing as a family over the empty context and that operations
act the same on families over the empty context as on contexts. Not including
an empty context changes the theory a bit. One example of a difference is that
if we want weakening by a context we shall have to explicitly state it as it 
would not follow from weakening by a family. A second example is that without
an empty context, some rule schemes (like the compatibility of extension with
the empty families, stated in \autoref{comp-0e} below) become slightly
assymetrical. Another example is that it would
not be possible to consider closed types and terms without an empty context.
Although we have opted to not include an empty context by default, we
shall give the rules in \autoref{empty-context} so that one can look them up
if they are desired.

{\color{red}Explain why empty families are needed}

The empty family over a context $\Gamma$ is introduced by the following rule
inference rule:
\begin{equation}
\inference
  { \jctx{\Gamma}
    }
  { \jfam{\Gamma}{\emptyf[\Gamma]}
    }
\end{equation}
Since it is not really necessary to annotate the empty family with its context,
we shall usually denote $\emptyf[\Gamma]$ by $\emptyf$.

The reader may wonder whether the empty family $\emptyf$ always has a
term. This shall follow from the rules stating the compatibility of extension
with the empty families in \autoref{comp-0e} below and from
identity terms (\autoref{identityterms}).
\begin{comment}
When contexts are viewed as statements, the empty context is the statement that
asserts nothing. Likewise, in any context one can assert nothing and this is
achieved by the empty family. 
Since somebody who states nothing asserts it all right, there is a term of the empty
family over any context. Moreover, any two terms of the empty family are always
judgmentally equal.
\begin{align}
& \inference
  { }
  { \jctx{\emptyc}
    }
  \\
& \inference
  { \jctx{\Gamma}
    }
  { \jfam{\Gamma}{\emptyf[\Gamma]}
    }
  \\
& \inference
  { \jctx{\Gamma}
    }
  { \jterm{\Gamma}{\emptyf[\Gamma]}{\emptytm[\Gamma]}
    }
  \\
& \inference
  { \jterm{\Gamma}{\emptyf[\Gamma]}{x}
    }
  { \jtermeq
      {\Gamma}
      {\emptyf[\Gamma]}
      {x}
      {\emptytm[\Gamma]}
    }
  \label{comp-00-t}
\end{align}

\label{comp-00}
Because every context is a family of contexts over the empty context, 
the empty context $\emptyc$ may be considered as a family of contexts over the empty
context. When we do this, we get $\emptyf[\emptyc]$.
\begin{equation}\label{eq:comp-00}
\inference
  { }
  { \jfameq
      {\emptyc}
      {\emptyc}
      {\emptyf[\emptyc]}
    }
\end{equation}
In the future, we shall denote $\emptyf[\Gamma]$ by $\emptyf$. The above rule
guarantees that this will not cause confusion. Likewise, we shall denote
$\emptytm[\Gamma]$ by $\emptytm$.
\end{comment}

In the following set of inference rules we state that the empty context and
the empty family are neutral objects when extension is concerned.
\label{comp-e0}\label{comp-0e}
\begin{align}
& \inference
  { \jctx{\Gamma}
    }
  { \jctxeq{\ctxext{\Gamma}{\emptyf}}{\Gamma}
    }
  \label{comp-e0-c}\\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{\ctxext{\emptyf}{A}}{A}
    }
  \label{comp-0e-f}
  \\
& \inference
  { \jfam{\Gamma}{A}
    }
  { \jfameq{\Gamma}{{A}{\emptyf}}{A}
    }
  \label{comp-e0-f}
%& \inference
%  { \jctx{\Gamma}
%    }
%  { \jctxeq{\ctxext{\emptyc}{\Gamma}}{\Gamma}
%    }
%  \label{comp-0e-c}\\
\end{align}
%Note the special case where we take $\Gamma$ to be the empty context.

@@

\subsection{Further judgmental equalities of the theory of contexts, families and
terms}
We have already seen that empty families, extension, weakening, substitution
and the identity terms are compatible with each other. There are, however, a
few more judgmental equalities explaining what happens when
\begin{enumerate}
\item something is weakend by the empty family. This is used to consider a term
      $x$ of $A$ in context $\Gamma$ as a morphism from $\emptyf$ to $A$ in
      context $\Gamma$;
\item something is substituted by the term of the empty family;
\item something is weakened by an extended family. This is Currying for weakening;
\item something is first weakened by a family $A$ in context $\Gamma$ and then
      substituted by a term of $A$. This is used to show that composition is
      compatible with itself (i.e. that composition is associative);
\item the identity terms are joining the party. Identity terms are empty contexts
      in certain models, so they should behave accordingly.
\end{enumerate}
In other words, the rules seem to not fit in the scheme that all operations are
compatible with each other, but in fact each of these rules plays a role in
completing the scheme (work out further).
